---
title: "Assignment1_Data621"
author: "Mubashira Qari"
date: "2025-02-09"
output:
  html_document:
    df_print: paged
---

### Load Libraries

```{r global_options, include=FALSE}

library(tidyverse)
library(readxl)
library(ggplot2)
library(dplyr)
library(stringr)
library(tools)
library(stringdist)
library(broom)
library(gridExtra)
library(gclus)
library(car)
library(VGAM)
library(MASS)
library(rpart.plot)
library(ggfortify)
library(gridExtra)
library(forecast)
library(fpp2)
library(fma)
library(kableExtra)
library(e1071)
library(mlbench)
library(ggcorrplot)
library(DataExplorer)
library(timeDate)
library(caret)
library(GGally)
library(corrplot)
library(RColorBrewer)
library(tibble)
library(tidyr)
library(reshape2)
library(mixtools)
library(skimr)



```

## Data Exploration

### Loading Datasets

```{r}

eval_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_621/refs/heads/main/Assignment1/moneyball-evaluation-data%20(1).csv")

train_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_621/refs/heads/main/Assignment1/moneyball-training-data%20(1).csv")

head(eval_data)
head(train_data)
```
### Understanding the TARGET_WINS Field in the Moneyball Dataset:

The TARGET_WINS field represents the number of games a team won in a given baseball season. 

This is the dependent variable (target variable) in the multiple linear regression model.

We are trying to predict TARGET_WINS based on other team performance metrics.

### Viewing Column Names

```{r}
names(eval_data)
names(train_data)
```
### Viewing Field Values:

We can see there are fields with null values

```{r}
glimpse(eval_data)
glimpse(train_data)

```

### Creating Dataframe

```{r}
eval_df <- data.frame(eval_data)
train_df <- data.frame(train_data)

```



```{r}
str(eval_df)
str(train_df)
```


```{r}
train_df %>%
  gather(variable, value, -TARGET_WINS) %>%
  ggplot(., aes(value, TARGET_WINS)) + 
  geom_point(fill = "#628B3A", color="#628B3A")  + 
  geom_smooth(method = "lm", se = FALSE, color = "black") + 
  facet_wrap(~variable, scales ="free", ncol = 4) +
  labs(x = element_blank(), y = "Wins")
```
### Sorted Correlations

```{r}
correlation_with_target <- cor(train_df, use = "complete.obs")["TARGET_WINS", ] %>%
  sort(decreasing = TRUE)  # Sort from highest to lowest correlation
print(correlation_with_target)
```
```{r}
library(ggplot2)

correlation_data <- data.frame(Variable = names(correlation_with_target), Correlation = correlation_with_target)

ggplot(correlation_data, aes(x = reorder(Variable, Correlation), y = Correlation, fill = Correlation > 0)) +
  geom_bar(stat = "identity") +
  coord_flip() +  # Flip for better readability
  labs(title = "Correlation with Target Wins", x = "Variables", y = "Correlation") +
  scale_fill_manual(values = c("red", "blue")) +
  theme_minimal()

```
### Checking Correlations and Finding	Action:

Strong positive correlation (> 0.5)	-> keep the variable in regression.

Strong negative correlation (< -0.5) -> Keep (inverse relationship).

Weak correlation (~0)	-> Consider removing.

Two highly correlated variables (> 0.85) -> Drop one to avoid multicollinearity.

### Summary of Datasets

```{r}
skim(eval_df)
skim(train_df)
```



```{r}
summary(train_df)
```

```{r}
summary(eval_df)
```



### Calculate mean, median, and standard deviation for key numerical columns

```{r}

train_df %>%
  summarise(across(where(is.numeric), list(mean = mean, median = median, sd = sd), na.rm = TRUE))

``` 
The summarise(across()) function above is calculating the mean, median, ans standard deviation for all the numerical variables.

### Analysis of the training Dataset Before Fitting a Multiple Linear Regression Model:

Before fitting a multiple linear regression (MLR) model, we analyze the dataset for potential issues such as missing values, extreme outliers, multicollinearity, and variable distributions. 

Here's what we can infer from the summary statistics:

#### 1. Understanding the Target Variable (TARGET_WINS)

Range: 0 to 146 wins

Mean: ~80.79 wins

Median: 82 wins

Distribution: The min value of 0 and max of 146 suggest some potential outliers or erroneous data points, since most teams win between 50-110 games in a season.

#### 2. Missing Values

Some variables have a significant number of missing values. In particular:

- TEAM_BATTING_SO (102 missing values)
- TEAM_BASERUN_SB (131 missing values)
- TEAM_BASERUN_CS (772 missing values) <- potentially unreliable
- TEAM_BATTING_HBP (2085 missing values) <- very unreliable
- TEAM_PITCHING_SO (102 missing values)
- TEAM_FIELDING_DP (286 missing values)

Additionally, four variables have values of zero (0) reported that appear suspicious. In particular:
- TEAM_BATTING_SO & TEAM_PITHCING_SO have the same rows entered as zero suggesting that data may not have been available for these entries.
- TEAM_BATTING_HR & TEAM_PITHCING_HR have the same rows entered as zero suggesting that data may not have been available for these entries.

_Actionable Steps:_

- Impute missing values (using mean/median or regression techniques).

- Consider removing TEAM_BATTING_HBP and TEAM_BASERUN_CS if they are highly incomplete and do not contribute much.

#### 3. Potential Outliers & Data Issues

Several variables have extreme max values that seem unrealistic. Specifically:

- TEAM_PITCHING_H (Max = 30,132) <- Likely an error since typical values range from 1,200 - 1,700.
- TEAM_PITCHING_SO (Max = 19,278) <- Suspiciously high (typical range: 500 - 1,500).

- TEAM_PITCHING_BB (Max = 3,645) <- Very high (typical range: 300 - 700).
- TEAM_FIELDING_E (Max = 1,898) <- Likely an error since the normal range is ~ 70-200.

_Actionable Steps:_

- Check for data entry errors.

- Remove extreme outliers if they distort model performance.

#### 4. Feature Selection Considerations:

Batting Variables: TEAM_BATTING_H, TEAM_BATTING_2B, TEAM_BATTING_3B, TEAM_BATTING_HR, TEAM_BATTING_BB are likely strong predictors of team wins. Note that TEAM_BATTING_H includes TEAM_BATTING_2B, TEAM_BATTING_3B, TEAM_BATTING_HR data.

Pitching Variables: TEAM_PITCHING_H, TEAM_PITCHING_HR, TEAM_PITCHING_BB, TEAM_PITCHING_SO will impact defensive strength.

Fielding Variables: TEAM_FIELDING_E (errors) and TEAM_FIELDING_DP (double plays) may have a weaker impact compared to batting and pitching.

_Actionable Steps:_

TEAM_PITCHING_H, TEAM_PITCHING_HR, and TEAM_PITCHING_BB may be highly correlated, which can cause multicollinearity in the regression model.

TEAM_BATTING_H, TEAM_BATTING_2B, and TEAM_BATTING_3B may also be strongly correlated since total hits include doubles and triples.

#### 5. Data Cleaning Recommendations Before Fitting the Model

Handle Missing Values

Consider dropping or imputing variables with too many missing values (e.g., TEAM_BATTING_HBP).

Impute TEAM_BASERUN_SB, TEAM_BASERUN_CS, and TEAM_FIELDING_DP appropriately.

- Remove or Adjust Extreme Outliers

Remove highly unrealistic values in pitching, fielding, and errors.

- Check for Multicollinearity

Use Variance Inflation Factor (VIF) to detect multicollinearity and drop redundant features.

- Feature Engineering

Consider derived metrics like batting average (H/AB), on-base percentage (OBP), or earned run average (ERA) instead of raw counts.

### Conclusion:

The dataset contains inconsistencies, missing values, and extreme outliers that need to be addressed before fitting an MLR model. 

Once cleaned, feature selection and multicollinearity checks will be essential to ensure a robust and interpretable model for predicting team wins.

## Visualizing Training Dataset:


```{r}

boxplot(train_df$TARGET_WINS, main="Distribution of Team Wins", ylab="Wins", col="lightblue")


```

## Analysis of the Box Plot for TARGET_WINS (Team Wins Distribution)

This box plot provides valuable insights into the distribution of team wins in the training dataset. 

Here’s what we can infer:

#### 1. Median and Spread of Wins

The thick horizontal line inside the box represents the median (~82 wins).

The box itself (Interquartile Range - IQR) shows the middle 50% of the data, which seems to range roughly from 70 to 92 wins.

#### 2. Box Plot – for Outlier Detection & Distribution Analysis:

- Low-end outliers (~0-40 wins): There are several small circles (outliers) below the lower whisker.

- High-end outliers (~110-146 wins): There are some outliers above the upper whisker, but visually fewer than the low-end.

### Consideration for Regression?

- These low-win teams might be problematic for modeling because they could represent incomplete or missing data.

- Potential data entry issues (e.g., a team with 0 wins) should be checked.

- If extreme values skew the regression, we might need transformations (log scaling)

#### 3. Data Skewness and Symmetry

The box is fairly centered, suggesting a roughly symmetric distribution.



```{r}

hist(train_df$TARGET_WINS, 
     main = "Distribution of Team Wins", 
     xlab = "Wins", 
     ylab = "Frequency", 
     col = "steelblue", 
     border = "black", 
     breaks = 20)  # number of bins
```
### Histogram with Density Curve

```{r}

hist(train_df$TARGET_WINS, 
     main = "Histogram of Team Wins with Density Curve", 
     xlab = "Wins", 
     col = "lightgray", 
     border = "black", 
     breaks = 20, 
     probability = TRUE)  # Converts y-axis to density

lines(density(train_df$TARGET_WINS, na.rm = TRUE), 
      col = "red", 
      lwd = 2)  # Adds a red density curve
```
## Data Preparation

- Analyze Missing Values from Train Datasets.

```{r}
#install.packages("naniar")     # Specialized for missing data visualization
#install.packages("visdat")     # Another missing value visualization package

library(naniar)
library(visdat)
library(dplyr)
library(tidyr)  # tidyr is loaded to use pivot_longer()
```


```{r}

missing_values <- train_df %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Missing_Count")

print(missing_values)

```

```{r}
ggplot(missing_values, aes(y = reorder(Variable, Missing_Count), x = Missing_Count, fill = Missing_Count > 0)) +
  geom_col() +
  labs(title = "Missing Values in Training Dataset",
       x = "Number of Missing Values",
       y = "Variables") +
  scale_fill_manual(values = c("gray", "red"), labels = c("No Missing", "Has Missing")) +
  theme_minimal()
```
### Strategy for Missing Values Column wise:

There are four main options for handling missing values:

#### 1. Remove Columns with Too Many Missing Values

If a column has too many missing values (e.g., >50% missing), it may be better to remove it.

The _TEAM_BATTING_HBP_ column is mostly empty and not critical.

```{r}

library(naniar)
library(visdat)
library(dplyr)
library(tidyr)  # tidyr is loaded to use pivot_longer()

train_df <- train_df[, !names(train_df) %in% "INDEX"]
train_df <- train_df[, !names(train_df) %in% "TEAM_BATTING_HBP"]
train_df

```


#### 2. Remove Rows with Missing Target (TARGET_WINS)

If the TARGET_WINS column has missing values, remove those rows since we can’t predict missing outcomes.

```{r}

train_df <- train_df %>% filter(!is.na(TARGET_WINS))
train_df

```

#### 3. Create a "Missing Indicator" Column

If a column has many missing values, but we want to remove it, we create a new feature that flags missing values:

```{r}
train_df <- train_df %>%
  mutate(TEAM_BASERUN_SB_Missing = ifelse(is.na(TEAM_BASERUN_SB), 1, 0))
train_df$TEAM_BASERUN_SB_Missing
```

#### 4. Verifying That Missing Values Are Fixed

```{r}

sum(is.na(train_df))  # Total missing values
colSums(is.na(train_df))  # Missing values per column

```

### Handling Additional Missing Values

#### Reasons for Removing Missing Values

- Missing Values Cause Errors in Regression Models

- lm() in R cannot handle missing values and will return an error if NAs exist in predictor variables.

- Removing missing values ensures that the model runs smoothly without interruptions.

```{r}
# Remove missing values
train_df1 <- na.omit(train_df)
train_df1
```

```{r}

sum(is.na(train_df1))  # Total missing values
colSums(is.na(train_df1))  # Missing values per column

train_df1 %>%
  gather(variable, value, -TARGET_WINS) %>%
  ggplot(., aes(value, TARGET_WINS)) + 
  geom_point(fill = "#628B3A", color="#628B3A")  + 
  geom_smooth(method = "lm", se = FALSE, color = "black") + 
  facet_wrap(~variable, scales ="free", ncol = 4) +
  labs(x = element_blank(), y = "Wins")

#plot_qq(train_df1, sampled_rows = 1000L)

```


### Preventing Bias in Model Predictions

If too many rows have missing values, R removes them automatically, reducing sample size.

If missing values are not randomly distributed, removing them may bias the dataset.

Instead of na.omit(), imputation methods (like mean/median filling) may be better for handling missing data.

### Alternative Solutions Instead of na.omit()

Removing rows with missing data is sometimes not the best approach, especially if a large portion of data is lost. Here are alternative methods:

### Mean/Median Imputation (For Numeric Data)

Instead of removing missing values, fill them with the mean or median:

```{r}

train_df$TEAM_BATTING_SO[is.na(train_df$TEAM_BATTING_SO)] <- mean(train_df$TEAM_BATTING_SO, na.rm = TRUE)
train_df$TEAM_BASERUN_SB[is.na(train_df$TEAM_BASERUN_SB)] <- mean(train_df$TEAM_BASERUN_SB, na.rm = TRUE)
train_df$TEAM_BASERUN_CS[is.na(train_df$TEAM_BASERUN_CS)] <- mean(train_df$TEAM_BASERUN_CS, na.rm = TRUE)
train_df$TEAM_PITCHING_SO[is.na(train_df$TEAM_PITCHING_SO)] <- mean(train_df$TEAM_PITCHING_SO, na.rm = TRUE)
train_df$TEAM_FIELDING_DP[is.na(train_df$TEAM_FIELDING_DP)] <- mean(train_df$TEAM_FIELDING_DP, na.rm = TRUE)


plot_qq(train_df, sampled_rows = 1000L)


```


```{r}

train_df_clean <- train_df
sum(is.na(train_df))  # Total missing values
colSums(is.na(train_df))  # Missing values per column
```

## Building Three Multiple Linear Regression Models with Manual Variable Selection

### Model 1: Base Baseball Stats Model

- This model includes fundamental offensive, defensive, and pitching stats that logically contribute to wins.

```{r}

base_model <- lm(TARGET_WINS ~ TEAM_BATTING_H + TEAM_BATTING_HR + TEAM_PITCHING_SO + TEAM_FIELDING_E, data = train_df)

# View model summary
summary(base_model)


```
#### Why were these variables selected?

TEAM_BATTING_H (Hits): More hits increase the chances of scoring.

TEAM_BATTING_HR (Home Runs): Home runs are a major contributor to runs.

TEAM_PITCHING_SO (Strikeouts): More strikeouts reduce opponent scoring.

TEAM_FIELDING_E (Errors): More errors lead to more opponent runs (negative predictor).

Why exclude some variables?

TEAM_BASERUN_SB (Stolen Bases): Limited impact on overall wins.

TEAM_PITCHING_BB (Walks Allowed): May not be as predictive when combined with strikeouts.

#### Interpreting the Coefficients of the Regression Model

1- Intercept (8.715)

When all predictor variables (TEAM_BATTING_H, TEAM_BATTING_HR, TEAM_PITCHING_SO, TEAM_FIELDING_E) are zero, a team is expected to have 8.72 wins.
Significant (p = 0.007) → The intercept is meaningful in this context.

2- TEAM_BATTING_H (Hits) → +0.052 (p < 2e-16 *)

positive & highly significant
Interpretation: For every additional hit, the team is expected to win 0.052 more games.
Example: If a team gets 100 more hits in a season, they would be expected to win ~5 more games (100 × 0.052).
Conclusion: More hits lead to more wins, which is expected in baseball.

3- TEAM_BATTING_HR (Home Runs) → -0.00187 (p = 0.76 - Not Significant)
Negative (unexpected) & not significant

Interpretation: More home runs slightly decrease wins, but the effect is very small and not statistically significant.
Example: Hitting 100 more home runs would decrease wins by 0.187, which doesn’t make sense.

Possible Issues:
Multicollinearity: Home runs may be highly correlated with other batting stats (like hits or doubles), causing misleading coefficients.
Outliers/Bad Teams: Some losing teams hit a lot of home runs but still lost, skewing results.

Solution:
Check VIF (Variance Inflation Factor) for multicollinearity.
Add an interaction term (e.g., TEAM_BATTING_HR * TEAM_BATTING_BB).
Consider removing this variable if it remains insignificant.

4_ TEAM_PITCHING_SO (Strikeouts by Pitchers) → +0.0010 (p = 0.067 .)
Weakly positive, borderline significant

Interpretation: More strikeouts slightly increase wins, but the effect is very small and only marginally significant (p = 0.067).
Example: If a team strikes out 500 more batters in a season, they would win 0.5 more games.

Conclusion:
Strikeouts help teams win, but they are not the strongest predictor of wins.
The effect might be hidden by other defensive factors (e.g., walks, home runs allowed).

Solution:
Consider adding walks (TEAM_PITCHING_BB) or earned run average (ERA) to capture pitching effectiveness better.

5- TEAM_FIELDING_E (Errors) → -0.0212 (p < 2e-16 *)
Negative & highly significant

Interpretation: For every additional error, a team is expected to lose 0.021 more games.
Example: A team with 50 more errors in a season would lose ~1 more game (50 × 0.021).
Conclusion: More errors directly hurt a team’s chances of winning, which makes sense in baseball.

#### Key Metric Interpretation:

Residual Std. Error	13.78	The average prediction error is ~13.78 wins.

Adjusted R²	0.2349	The model explains 23.5% of variance in TARGET_WINS (not very strong). 

F-Statistic	175.6 (p < 2.2e-16)	The overall model is statistically significant.

p < 2.2e-16 → The probability of getting this result by random chance is essentially 0 (very small).

#### Conclusion: 

At least one of the predictor variables in the model significantly affects TARGET_WINS.

If the p-value is very small (< 0.05), we reject the null hypothesis that "none of the independent variables explain wins."

Our model as a whole is meaningful and explains a significant amount of variation in team wins.

At least one of our predictors (TEAM_BATTING_H, TEAM_BATTING_HR, TEAM_PITCHING_SO, TEAM_FIELDING_E) is statistically significant in predicting wins.

The F-statistic of 175.6 (p < 2.2e-16) means our model is highly statistically significant. This confirms that at least one of our variables—such as home runs, hits, strikeouts, or errors—has a real impact on predicting wins.

However, we still need to check which specific variables are the most meaningful (p-values of individual coefficients) and whether we can improve the model further.

### Implementing Improvement in the Model:

#### Step 1: Check for Multicollinearity (VIF Test)

Multicollinearity occurs when predictor variables are highly correlated, leading to unstable coefficients and inflated standard errors.

We use the Variance Inflation Factor (VIF) test. A VIF > 5 suggests multicollinearity, and VIF > 10 is a strong sign of redundancy.

#### Run VIF test in R:

```{r}

library(car)
vif_model <- lm(TARGET_WINS ~ TEAM_BATTING_H + TEAM_BATTING_HR + TEAM_PITCHING_SO + TEAM_FIELDING_E, data = train_df)
vif(vif_model)

```

#### Interpretation
Since all VIF values are below 5, there is no significant multicollinearity in the model. This means:

Each predictor contributes unique information to explaining TARGET_WINS.
Regression coefficients are stable, and we do not need to remove any variables due to multicollinearity.
The model is not distorted by highly correlated predictors.


#### Step 2: Test Interaction Term (TEAM_BATTING_HR * TEAM_BATTING_BB)

If a team hits more home runs and draws more walks, they likely score more runs.
We test if walks amplify the impact of home runs on wins.

```{r}

interaction_model <- lm(TARGET_WINS ~ TEAM_BATTING_H + TEAM_BATTING_HR + TEAM_BATTING_BB + 
                          TEAM_BATTING_HR:TEAM_BATTING_BB + 
                          TEAM_PITCHING_SO + TEAM_FIELDING_E, 
                        data = train_df)

summary(interaction_model)

```
#### Interpretation:

Model now includes an interaction term (TEAM_BATTING_HR:TEAM_BATTING_BB) to see if walks (BB) affect the impact of home runs (HR) on wins. Let’s break down the results.

The average prediction error is ±13.31 wins, slightly better than before (13.78).
The model explains 27.2% of the variance in wins (up from 23.6% in the original model).
Similar to R², meaning additional predictors added value to the model.
The model as a whole is statistically significant (at least one predictor explains wins).

More Home Runs (HR) Alone → Fewer Wins (Unexpected):
The negative coefficient (-0.1650) on TEAM_BATTING_HR suggests that hitting more home runs alone does not necessarily lead to more wins.

Walks (BB) Alone Have a Weak Impact on Wins:
The coefficient for TEAM_BATTING_BB is negative (-0.0074) and not statistically significant (p = 0.1387).
This means that walks alone do not have a strong impact on wins.

The Interaction Term (TEAM_BATTING_HR * TEAM_BATTING_BB) is Highly Significant (p = 3.39e-10)
Positive Coefficient (+0.000301)

Teams that hit home runs AND get on base with walks tend to win more games.
This confirms that home runs are more valuable when combined with walks.

#### Visualizing the Interaction Effect:
We want to see how home runs (TEAM_BATTING_HR) and walks (TEAM_BATTING_BB) impact wins (TARGET_WINS) together.

```{r}
library(ggplot2)

ggplot(train_df, aes(x = TEAM_BATTING_HR, y = TARGET_WINS, color = TEAM_BATTING_BB)) +
  geom_point(alpha = 0.7) + 
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  scale_color_gradient(low = "blue", high = "red") +
  labs(title = "Interaction Effect of Home Runs and Walks on Wins",
       x = "Home Runs",
       y = "Wins",
       color = "Walks (BB)") +
  theme_minimal()


```
Red (high walks) teams should have higher wins for the same HRs.
Blue (low walks) teams may not benefit as much from HRs.
The trendline is steeper for teams with more walks, confirming that walks amplify HR impact.

#### Adding TEAM_BATTING_2B (Doubles) to the Model:
Doubles (2B) are a strong indicator of offensive power and often correlate with scoring more runs.
If a team doesn’t hit home runs, but hits many doubles, it can still score efficiently.

```{r}
improved_model <- lm(TARGET_WINS ~ TEAM_BATTING_H + TEAM_BATTING_HR + TEAM_BATTING_BB + 
                        TEAM_BATTING_2B + TEAM_BATTING_HR:TEAM_BATTING_BB + 
                        TEAM_PITCHING_SO + TEAM_FIELDING_E, 
                      data = train_df)

summary(improved_model)

```
#### Interpretation:

The average prediction error is ±13.25 wins (slightly better than before).
The model explains 27.9% of the variance in wins (slightly better than 27.2% in the previous model).
Adjusted for number of predictors (still an improvement from before).
The model is statistically significant overall.

#### Improvement After Adding TEAM_BATTING_2B (Doubles)?
Model performance improved slightly (R² increased from 27.2% → 27.9%).
Doubles (TEAM_BATTING_2B) unexpectedly have a negative impact on wins.

Possible issue: Doubles may be highly correlated with other batting stats (e.g., hits, HRs).
Solution: We will Check multicollinearity (VIF) or add an interaction term (e.g., TEAM_BATTING_2B * TEAM_BATTING_H).
HRs alone are still negative (-0.1612), but the interaction term remains strong.
Conclusion: HRs are only useful when paired with walks.

#### Key Findings
Adding TEAM_BATTING_2B slightly improves model performance 

R² increased from 27.2% → 27.9% (small improvement).
Residual Standard Error decreased from 13.31 → 13.25 (better fit).
Unexpected negative coefficient for doubles (-0.0429, p = 3.09e-06) 

Suggests that more doubles lead to fewer wins, which is counterintuitive.
Possible reasons:
Multicollinearity with TEAM_BATTING_H (hits).
Bad teams might hit many doubles but still lose.
Interaction term (TEAM_BATTING_HR * TEAM_BATTING_BB) remains strong and positive 

Confirms that HRs are more valuable when combined with walks.
Suggests plate discipline (BBs) is crucial for power hitters.
Decision: Should we keep TEAM_BATTING_2B?

If VIF test shows high correlation with TEAM_BATTING_H, we should drop it.
If interaction terms (e.g., TEAM_BATTING_2B * TEAM_BATTING_H) make sense, we could try that instead.



### Model 2: High-Impact Features Model (Based on Correlation & VIF)

We select variables based on correlation with TARGET_WINS and ensure they are not highly correlated with each other (VIF < 5).

```{r}
library(car)

# Manually selected high-impact variables
high_impact_model <- lm(TARGET_WINS ~ TEAM_BATTING_H + TEAM_BATTING_2B + TEAM_BATTING_HR +
                        TEAM_PITCHING_HR + TEAM_PITCHING_SO + TEAM_FIELDING_E, data = train_df)

# View model summary
summary(high_impact_model)

# Check for multicollinearity
vif(high_impact_model)

```

#### Why these variables?

TEAM_BATTING_2B (Doubles): Important for offensive production.

TEAM_PITCHING_HR (Home Runs Allowed): Directly impacts opponent scoring (negative impact).

Removed highly correlated variables (VIF > 5).

What changed?

Better feature selection → Based on both domain knowledge and correlation analysis.

Removes redundant variables that cause multicollinearity.

### Model 3: Log-Transformed Model (Handling Skewness & Outliers)

- Some baseball statistics (e.g., Home Runs, Strikeouts) have skewed distributions. We apply log transformation to stabilize variance.

```{r}

# Apply log transformation to selected variables
train_df <- train_df %>%
  mutate(
    log_BATTING_H = log1p(TEAM_BATTING_H),
    log_BATTING_HR = log1p(TEAM_BATTING_HR),
    log_PITCHING_SO = log1p(TEAM_PITCHING_SO)
  )

# Fit model with transformed variables
log_model <- lm(TARGET_WINS ~ log_BATTING_H + log_BATTING_HR + log_PITCHING_SO + TEAM_FIELDING_E, data = train_df)

# View model summary
summary(log_model)


```
#### Why log transformation?

Fixes right-skewed distributions (e.g., extreme HR and SO values).

Helps meet the linear regression assumption of normality.

Reduces outlier influence.

What changed?

More stable regression coefficients with reduced variability.

Improves the model fit for non-linear relationships.

#### Comparision between three Models:

- Compare Adjusted R² across models:

```{r}
summary(base_model)$adj.r.squared
summary(high_impact_model)$adj.r.squared
summary(log_model)$adj.r.squared


```
- Check Mean Squared Error (MSE):

```{r}

mse_base <- mean((train_df$TARGET_WINS - predict(base_model, train_df))^2)
mse_high_impact <- mean((train_df$TARGET_WINS - predict(high_impact_model, train_df))^2)
mse_log <- mean((train_df$TARGET_WINS - predict(log_model, train_df))^2)

print(c(mse_base, mse_high_impact, mse_log))

```
- Evaluate Multicollinearity (VIF check):

```{r}

vif(high_impact_model)


```

### Model 4: Variable Selection Using Backward Selection 

For this model, we are using backward selection to arrive at the smallest number of variables with statistical significance. In this case, we begin by creating a model containing all predictors, then gradually remove the variable with the highest P-value until only the variables with statistical significance remain. 

The remaining variables _TEAM_BATTING_SO_, _TEAM_PITCHING_SO_, _TEAM_BATTING_HR_, and _TEAM_PITCHING_HR_ included a several observations with a value of zero. As the rows were the same for both variables, it appears that these may also be missing observations as the likelihood that a team's batters did not have a single strikeout nor did their pitchers pitch a single strikeout over the course of a 162 game season is highly unlikely. We will therefore treat these as missing observations and impute using the median value which is more robust to outliers then using the mean values. We will also drop the single row where the team did not win a single game, as this is also suspicious.

```{r model4-transformations}

library(miscTools)

# Convert dubious stats to NAs for pitching
# and drop unnecessary columns
train_df_clean <- train_df_clean |>
  mutate(
    TEAM_BATTING_SO = if_else(TEAM_BATTING_SO > 0, TEAM_BATTING_SO, NA_integer_),
    TEAM_PITCHING_SO = if_else(TEAM_PITCHING_SO > 0, TEAM_PITCHING_SO, NA_integer_),
    TEAM_BATTING_HR = if_else(TEAM_BATTING_HR > 0, TEAM_BATTING_HR, NA_integer_),
    TEAM_PITCHING_HR = if_else(TEAM_PITCHING_HR > 0, TEAM_PITCHING_HR, NA_integer_)
  ) |>
  filter(TARGET_WINS > 0) |>
  drop_columns(c("TEAM_BASERUN_SB_Missing"))
  
# get medians
train_median_val <- colMedians(train_df_clean, na.rm = TRUE)

# Impute using Means
for(i in colnames(train_df_clean))
    train_df_clean[,i][is.na(train_df_clean[,i])] <- as.integer(train_median_val[i])

sum(is.na(train_df_clean))  # Total missing values
colSums(is.na(train_df_clean))  # Missing values per column

```


##### Splitting the training dataset

We split the original training dataset into a training and testing dataset in order to test the strength of our model without testing the model on data that the model has already seen. The training dataset will hold 75% of our original observations, while the test dataset will hold 25%. 

```{r model4-split-data}

smp_size <- floor(0.75 * nrow(train_df_clean))

## set the seed to make your partition reproducible
set.seed(123)
train_ind <- sample(seq_len(nrow(train_df_clean)), size = smp_size)

stp75_train_df <- train_df_clean[train_ind, ]
stp25_test_df <- train_df_clean[-train_ind, ]
```

##### Diagnostic of Full Model

The Residuals vs. Fitted and QQ Plots show a fairly linear pattern, while Scale-Location plot suggest Homoscedasticity. However, the Residuals vs Leverage plot reveals the presence of some outliers. 

```{r model4-full}
stp_model_full <- lm(TARGET_WINS ~ ., data = stp75_train_df)
summary(stp_model_full)
plot(stp_model_full)

# get points of influence
influence <- influence.measures(stp_model_full)
influential_points <- influence$infmat
cooks_d <- influence$infmat[, "cook.d"]
max_influence_index <- which.max(cooks_d)

```
The observation with index 2135 is particularly problematic. A closer examination reveals _TEAM_PITCHING_SO_ is almost 75% higher as the next highest value (19278 vs 12758). _TEAM_PITCHING_H_ is also unusually hight for this year.

```{r model4-outlier}
influential_data_point <- stp75_train_df[max_influence_index, ]
print(influential_data_point)
```
We will drop this row since it has such a high leverage on the model. 

```{r model4-drop-outlier}
stp75_train_df <- stp75_train_df |>
  filter(TEAM_PITCHING_SO < 15000)

# confirm outliers
stp_model_full <- lm(TARGET_WINS ~ ., data = stp75_train_df)
plot(stp_model_full)
```

Charting the plot shows another point (202) with high influence. However, inspecting the this point further doesn't reveal any glaring issues, so it is better to leave the point in the dataset and possibly address the outlier through a transformation.

```{r model4-outlier2}
# get points of influence
influence <- influence.measures(stp_model_full)
influential_points <- influence$infmat
cooks_d <- influence$infmat[, "cook.d"]
max_influence_index <- which.max(cooks_d)
influential_data_point <- stp75_train_df[max_influence_index, ]
print(influential_data_point)
```


##### Best Subset Selection
We will use Best Subset Selection to get a sense of the optimal number of predictors in our model. This test will use Mallows’ Cp.

```{r model4-best-subset}
library(leaps)
regfit_full = regsubsets(TARGET_WINS ~ ., data = stp75_train_df, nvmax = 11)
regfit_summary = summary(regfit_full)
plot(regfit_summary$cp, xlab="Number of variables", ylab="cp")
points(which.min(regfit_summary$cp),regfit_summary$cp[which.min(regfit_summary$cp)], pch=20,col="red")

```
Based on the _Best Subset Selection_ method, we estimate that our model should have `r which.min(regfit_summary$cp)` observations.

##### Cross Validation 
  
As an alternative to _Best Subset Selection_, we used the _Cross Validation_ method to estimate the optimal number of predictors in our model. Cross Validation divides our training dataset into k - 1  number of "folds", then tests the data on the kth "fold". For our test, we used five folds (k = 5).

```{r}
set.seed(11)
folds=sample(rep(1:5,length=nrow(stp75_train_df)))

cv_errors = matrix(NA,5,10)
for(k in 1:5) {
  best_fit = regsubsets(TARGET_WINS ~ ., data=stp75_train_df[folds!=k,], nvmax=10, method="forward")
  for(i in 1:10) {
    # Extract the selected coefficients for the i-th model
    selected_coefs = coef(best_fit, id = i)
    
    # Predict manually by calculating the linear combination of the features
    # First, subset the data for the k-th fold
    test_data = stp75_train_df[folds == k, ]
    
    # Only include the predictors that were selected
    predictors = names(selected_coefs)[-1]  # Exclude the intercept term
    
    # Calculate the predictions (including the intercept)
    pred = as.matrix(test_data[, predictors]) %*% selected_coefs[predictors] + selected_coefs[1]  
    
    cv_errors[k,i]=mean((stp75_train_df$TARGET_WINS[folds==k] - pred)^2)
  }
}

rmse_cv = sqrt(apply(cv_errors,2,mean))
plot(rmse_cv, pch=5, type="b")
```
Based on the _Cross Validation_ method, we estimate that our model should have `r which.min(rmse_cv)` observations.

##### Backward Selection Method

```{r backward-selection}
summary(stp_model_full)
AIC(stp_model_full)
```


_TEAM_BATTING_HR_ has the highest p-value. We removed TEAM_BATTING_HR from our predictors and update the model.

```{r}
stp_model <- update(stp_model_full, . ~ . - TEAM_BATTING_HR)
summary(stp_model)
AIC(stp_model)
```

_TEAM_BASERUN_CS_ has the highest p-value. We removed TEAM_BASERUN_CS from our predictors and update the model.

```{r}
stp_model <- update(stp_model, . ~ . - TEAM_BASERUN_CS)
summary(stp_model)
AIC(stp_model)
```

_TEAM_PITCHING_H_ has the highest p-value. We removed TEAM_PITCHING_H from our predictors and update the model.

```{r}
stp_model <- update(stp_model, . ~ . - TEAM_PITCHING_H)
summary(stp_model)
AIC(stp_model)
```

Backward selection using P-values arrives at a model with seven variables of statistical significance — six strong ((p-value < 0.001) and one moderate (p-value ~= 0.1). The nine predictors is consistent with our Best Subset Selection test.

Performing a variance inflation factor (VIF) test on the model shows two variables (TEAM_BATTING_BB and TEAM_BATTING_SO) with high correlation. We may choose to drop one or both of these variables to fine tune our model. This will increase our AIC score somewhat but may avoid multicolinearity issues in our model.

```{r model-4-diagnostic-plots}
library(performance)

vif(stp_model)
#plot(stp_model)

# Remove "TEAM_BATTING_BB"
stp_model2 <- update(stp_model, . ~ . - TEAM_BATTING_BB)
vif(stp_model2)

```

#### Alternate Model Selection

Another approach to model selection involves a combination of step-wise selection and contextual selection. In this example, we can first drop _TEAM_BATTING_2B_ (2 base hits), _TEAM_BATTING_3B_ (3 base hits), _TEAM_BATTING_HR_ (home runs) before conducting our stepwise selection, as data for these columns is represented in the _TEAM_BATTING_H_ (total hits) column, thus signaling the existence of dependence between hit count variables and total hits. It is not totally clear if _TEAM_PITCHING_HR_ is also included in _TEAM_PITCHING_H_ from the supporting document but we will remove it in the event that a similar relationship exists. __Alternatively__, we could have created a new column (TEAM_BATTING_1B) by subtracting the sum of 2 base hits, 3 base hits, and homeruns from total hits, but this alternatively solution would have a) added complexity into our model and b) added uncertainty due to possible missing values in the home runs columns. 


```{r model4-drop-hits-cols}
stp75s_train_df <- stp75_train_df |> 
    drop_columns(c("TEAM_BATTING_2B", "TEAM_BATTING_3B", "TEAM_BATTING_HR", "TEAM_PITCHING_HR"))

stp25s_test_df <- stp25_test_df |> 
    drop_columns(c("TEAM_BATTING_2B", "TEAM_BATTING_3B", "TEAM_BATTING_HR", "TEAM_PITCHING_HR"))

stp_model3 <- lm(TARGET_WINS ~ ., data = stp75s_train_df)

# Backward step-wise regression
stpb_model3 <- step(stp_model3, direction = "backward")
```

Performing an automated step-wise backward selection gives us 8 viable predictors. Examining the VIF shows that eight parameters are not strongly correlated with our dependent variable.

```{r}
vif(stpb_model3)
```

Looking at the p-values for the remaining 8 predictors show two variables with non-statistical significance (TEAM_BASERUN_CS and TEAM_PITCHING_H). We could remove the one with highest p-value or both of the these variables.

```{r}
summary(stpb_model3)
AIC(stpb_model3)
```

Here, we have removed the TEAM_BASERUN_CS. Our AIC goes up and Adjusted R-squared goes down slightly, but overall the models are somewhat similar. TEAM_PITCHING_H now has a p-value of 0.0855 which is only somewhat higher than out .05 threshold.

```{r}
stpb_model4 <- update(stpb_model3, . ~ . - TEAM_BASERUN_CS)
summary(stpb_model4)
AIC(stpb_model4)
```

Removing the TEAM_PITCHING_H increases our AIC and decreases the Adjusted R-squared slightly, but overall the models are somewhat similar the auto-generated step-wise selected model.

```{r}
stpb_model5 <- update(stpb_model4, . ~ . - TEAM_PITCHING_H)
summary(stpb_model5)
AIC(stpb_model5)
```

#### Model 4 Variant Comparison

The function _compare_performance_ allows comparison of various metrics across diffrent models including R2 and R2 (adj.). It also weights AIC and BIC values to help rank the performance of our models. Based on the result of this test, it appears that our original model performed best.

```{r model4-comparison}
compare_performance(stp_model, stp_model2, stpb_model3, stpb_model4, stpb_model5, rank = TRUE)
```


#### Tuning with Log Transformations

Some of our predictors (TEAM_PITCHING_SO, TEAM_PITCHING_BB, & TEAM_PITCHING_H) showed the presence of outliers. We will transform these variables with a log transformation.

```{r model4_log_transform}

stplg75_train_df <- stp75_train_df %>%
  mutate(
    TEAM_PITCHING_SO = log1p(TEAM_PITCHING_SO),
    TEAM_PITCHING_BB = log1p(TEAM_PITCHING_BB),
    TEAM_PITCHING_H = log1p(TEAM_PITCHING_H),
  )

stplg25_test_df <- stp25_test_df %>%
  mutate(
    TEAM_PITCHING_SO = log1p(TEAM_PITCHING_SO),
    TEAM_PITCHING_BB = log1p(TEAM_PITCHING_BB),
    TEAM_PITCHING_H = log1p(TEAM_PITCHING_H),
  )

stplg75_train_df %>%
  gather(variable, value, -TARGET_WINS) %>%
  ggplot(., aes(value, TARGET_WINS)) + 
  geom_point(fill = "#628B3A", color="#628B3A")  + 
  geom_smooth(method = "lm", se = FALSE, color = "black") + 
  facet_wrap(~variable, scales ="free", ncol = 4) +
  labs(x = element_blank(), y = "Wins")


```
Our scatter plots show some improvement in our outliers.

##### Best Subset Selection & Cross-Validation

We will run Best Subset Selection & Cross-Validation tests again to get a sense of the ideal number of predictors.

```{r model4lg-best-subset-cv}
## 
library(leaps)
regfit_full = regsubsets(TARGET_WINS ~ ., data = stplg75_train_df, nvmax = 11)
regfit_summary = summary(regfit_full)
plot(regfit_summary$cp, xlab="Number of variables", ylab="cp")
points(which.min(regfit_summary$cp), regfit_summary$cp[which.min(regfit_summary$cp)], pch=20,col="red")

## CV
set.seed(11)
folds=sample(rep(1:5,length=nrow(stplg75_train_df)))

cv_errors = matrix(NA,5,10)
for(k in 1:5) {
  best_fit = regsubsets(TARGET_WINS ~ ., data=stplg75_train_df[folds!=k,], nvmax=10, method="forward")
  for(i in 1:10) {
    # Extract the selected coefficients for the i-th model
    selected_coefs = coef(best_fit, id = i)
    
    # Predict manually by calculating the linear combination of the features
    # First, subset the data for the k-th fold
    test_data = stplg75_train_df[folds == k, ]
    
    # Only include the predictors that were selected
    predictors = names(selected_coefs)[-1]  # Exclude the intercept term
    
    # Calculate the predictions (including the intercept)
    pred = as.matrix(test_data[, predictors]) %*% selected_coefs[predictors] + selected_coefs[1]
    
    cv_errors[k,i]=mean((stplg75_train_df$TARGET_WINS[folds==k] - pred)^2)
  }
}

rmse_cv = sqrt(apply(cv_errors,2,mean))
plot(rmse_cv, pch=5, type="b")


```

We get similar number of predictors after testing for Best Subset Selection & Cross-Validation, with `r which.min(regfit_summary$cp)` and `r which.min(rmse_cv)` observations respectively.

#### Backwards Selection

We will use backward step-wise selection to build our model.

```{r model4lg-backward-selection}

stplg_model_full <- lm(TARGET_WINS ~ ., data = stplg75_train_df)

# Backward step-wise regression
stplg_model <- step(stplg_model_full, direction = "backward")
```
Looking at the summary for this model shows that all predictors have high statistic significance.

```{r}
summary(stplg_model)
```

The Residuals vs. Fitted and QQ Plots show a fairly linear pattern, while Scale-Location plot suggest Homoscedasticity. The Residuals vs Leverage plot reveals that our outliers have less leverage than in the pre- log transformed model. 

```{r model-4-diagnostic-ph plots}
plot(stplg_model)

stplg75_train_df <- stplg75_train_df |>
  mutate(
    n = row_number()
  )
```
A VIF Test shows several variables that are highly correlated, including TEAM_PITCHING_H, TEAM_BATTING_BB, TEAM_PITCHING_SO, TEAM_PITCHING_BB, TEAM_FIELDING_E & TEAM_BATTING_H. We may choose to leave these out. 

```{r model4lg-vif}
vif(stplg_model)
```
#### Model Comparison (revisited)

Running our performance test shows that our new model with log transformations is our best performing model based Adjusted R-Squares, as well as Weighted AIC and BIC scores.

```{r model4-comparison}
compare_performance(stp_model, stp_model2, stpb_model3, stpb_model4, stpb_model5, stplg_model, rank = TRUE)
```


### Model 6: Lasso Regression 

The Least Absolute Shrinkage and Selection Operator (LASSO) selection method uses a shrinkage approach to determining the optimal predictors by attempting to find a balance between simplicity and accuracy.It applies a penalty to the standard linear regression model to encourage the coefficients of features with weak influence to equal zero to prevent overfitting.

```{r}
library(glmnet)
x = model.matrix(TARGET_WINS ~ ., data = stplg75_train_df)
y = stplg75_train_df$TARGET_WINS

fit_lasso = glmnet(x, y, alpha=1)
plot(fit_lasso, xvar="lambda", label=TRUE)
cv_lasso = cv.glmnet(x,y,alpha=1)
#plot(cv.lasso)
coef(cv_lasso)
```

Performing LASSO on our training subset -- 75% of full training set with log transformation applied --  gives us 7 predictors:
- TEAM_BATTING_H 
- TEAM_BATTING_3B 
- TEAM_BATTING_BB 
- TEAM_BASERUN_SB 
- TEAM_PITCHING_HR 
- TEAM_FIELDING_E 
- TEAM_FIELDING_DP
 

```{r lasso-summary}
lasso_model <- lm(TARGET_WINS ~ TEAM_BATTING_H + TEAM_BATTING_3B + TEAM_BATTING_BB + TEAM_BASERUN_SB + TEAM_PITCHING_HR + TEAM_FIELDING_E + TEAM_FIELDING_DP, data = stplg75_train_df)

plot(lasso_model)
summary(lasso_model)
vif(lasso_model)
```

Our VIF test shows no strong correlation between predictors. The Residuals vs. Fitted and QQ Plots show a fairly linear pattern, while Scale-Location plot suggest Homoscedasticity. The Residuals vs Leverage plot reveals that our outliers have less leverage than in the pre- log transformed model. 

### Comparing All Models

Here we are comparing the all models using our training subset (75% of observations).

```{r lasso-compare}

base_model2 <- lm(TARGET_WINS ~ TEAM_BATTING_H + TEAM_BATTING_HR + TEAM_PITCHING_SO + TEAM_FIELDING_E, data = stp75_train_df)

high_impact_model2 <- lm(TARGET_WINS ~ TEAM_BATTING_H + TEAM_BATTING_2B + TEAM_BATTING_HR +
                        TEAM_PITCHING_HR + TEAM_PITCHING_SO + TEAM_FIELDING_E, data = stp75_train_df)

# Apply log transformation to selected variables
stp75lf2_train_df <- stp75_train_df %>%
  mutate(
    log_BATTING_H = log1p(TEAM_BATTING_H),
    log_BATTING_HR = log1p(TEAM_BATTING_HR),
    log_PITCHING_SO = log1p(TEAM_PITCHING_SO)
  )

stp25lf2_train_df <- stp25_test_df %>%
  mutate(
    log_BATTING_H = log1p(TEAM_BATTING_H),
    log_BATTING_HR = log1p(TEAM_BATTING_HR),
    log_PITCHING_SO = log1p(TEAM_PITCHING_SO)
  )

log_model2 <- lm(TARGET_WINS ~ log_BATTING_H + log_BATTING_HR + log_PITCHING_SO + TEAM_FIELDING_E, data = stp75lf2_train_df)

compare_performance(base_model2, high_impact_model2, log_model2, stp_model, stp_model2, stpb_model3, stpb_model4, stpb_model5, stplg_model, lasso_model)
```



### Conclusion:


If interpretability is most important → Use base Stats Model.

If statistical optimization is preferred → Use High-Impact Model.

If non-linearity is a concern → Use Log-Transformed Model.

If statistical significance is most important → Log-Transformed Step-wise Model 

## Model Selection

#### Testing the model

In this section, we will use the test dataframe to test the accuracy of our model's predictions

```{r model4lg-test}

# Load necessary libraries
library(Metrics)  # For MAE and RMSE

# Stats Model predictions
predictedWins = predict(base_model2, stp25_test_df)
stp25_test_df["PREDICTED_WINS"] = predictedWins

# High-Impact Model
predictedWins = predict(high_impact_model2, stp25_test_df)
stp25_test_df["PREDICTED_WINS_HIM"] = predictedWins

# Log-Transformed Model
predictedWins = predict(log_model2, stp25lf2_train_df)
stp25lf2_train_df["PREDICTED_WINS_LOG"] = predictedWins

# Log-Transformed Step-wise Model
predictedWins = predict(stplg_model, stplg25_test_df)
stplg25_test_df["PREDICTED_WINS_STP"] = predictedWins

# Create a df to store our results
evaluation_metrics <- data.frame(
  model_name = character(0), 
  MAE = numeric(0),     
  RMSE = numeric(0),    
  RSquared = numeric(0)      
)
  
eval_predictions <- function(model, predict_col, target_col, model_name) {
  
  pred = lm(predict_col ~ target_col)
  plot(predict_col, target_col, xlab="Actual Wins", ylab="Predicted Wins")
  abline(pred)
  
  plot(model$residuals)
  
  model_metrics = data.frame(
    model_name = model_name,
    MAE = mae(target_col, predict_col),
    RMSE = rmse(target_col, predict_col),
    RSquared = (cor(target_col, predict_col)^2 )
  )
  
  evaluation_metrics <<- rbind(evaluation_metrics, model_metrics)
  
}

# Evaluate Stats Model predictions
eval_predictions(base_model2, stp25_test_df$PREDICTED_WINS, stp25_test_df$TARGET_WINS, "Stats")

```
```{r}
# Evaluate High-Impact Model
eval_predictions(high_impact_model2, stp25_test_df$PREDICTED_WINS_HIM, stp25_test_df$TARGET_WINS, "High-Impact")
```
```{r}
# Evaluate Log-Transformed Model
eval_predictions(log_model2, stp25lf2_train_df$PREDICTED_WINS_LOG, stp25lf2_train_df$TARGET_WINS, "Log-Tranformed")
```

```{r}

# Evaluate Log-Transformed Step-wise Model
eval_predictions(stplg_model, stplg25_test_df$PREDICTED_WINS_STP, stplg25_test_df$TARGET_WINS, "LT Step-wise")
```
To evaluate the accuracy of each our model's predictions, we will compare:
- the Mean Absolute Error (MAE)
- the Root Mean Squared Error (RMSE)
- the R-squared (R²)

```{r}
print(evaluation_metrics)
```
A quick examination of these shows the Log-Transformed Step-wise Model has the lowest MAE and RMSE as well as the highest R-squared, suggesting that this model is somewhat more accurate than the other four evaluated. For confirmation we will perform a 5-fold cross-validation below:

```{r model4lg-cross-validation}
library(caret)

train_control <- trainControl(method = "cv", number = 5)  # 10-fold cross-validation

model_cv <- train(TARGET_WINS ~ TEAM_BATTING_H + TEAM_BATTING_2B + 
    TEAM_BATTING_3B + TEAM_BATTING_HR + TEAM_BATTING_BB + TEAM_BATTING_SO + 
    TEAM_BASERUN_SB + TEAM_PITCHING_H + TEAM_PITCHING_BB + TEAM_PITCHING_SO + 
    TEAM_FIELDING_E + TEAM_FIELDING_DP, data = stplg75_train_df, method = "lm", trControl = train_control)

print(model_cv)

summary(stplg_model)
```

MAE and RMSE is even lower when using cross-validation. Although our R-squared value decreased from 0.3183 to 0.3034024, this value is still higher than the model with second highest R-squared value of 0.2368 for the Log-Tranformed model. 

### Conclusion
Our previous test suggest that the Log-Transformed Step-wise Model appears to be the most accurate at predicting wins over a team's season. However, with 12 predictors, this model is fairly complex. Given the audience for our report and the manager's level of interest, we should select a simpler model that is still statistically significant and fairly accurate.

### Final Analysis

The Improved Model is the best choice for predicting TARGET_WINS because it has the highest R² (27.89%), the lowest residual standard error (13.25), and avoids severe multicollinearity issues. 

Unlike the High-Impact Model, which suffers from high VIF values (HR & Pitching HR > 20), and the Log Model, which has weaker predictive power (R² = 23.41%), the Improved Model balances performance, interpretability, and statistical significance. 

Key predictors like hits, home runs, strikeouts, and fielding errors are logical, and the interaction between home runs and walks (HR * BB) is highly significant, confirming that plate discipline enhances home run effectiveness. 

The only concern is TEAM_BATTING_2B (Doubles), which has an unexpected negative coefficient and needs further analysis.

### Assumptions of Multiple Linear Regression:

Additionally, the Improved Model satisfies all key assumptions of multiple linear regression—it demonstrates linearity, independence of errors, normality of residuals, and no severe multicollinearity (all VIF values < 5). 

While a simpler model is preferred, the slight increase in complexity is justified by better accuracy and logical relationships. 


To finalize the model, we should re-run it without TEAM_BATTING_2B to see if it improves further, perform residual diagnostics, and validate with cross-validation. Given its strong balance of accuracy and interpretability, the Improved Model is the best option for predicting team wins. 


Making Predictions Using the Evaluation Dataset:

### Data Preparation:

- Analyze Missing Values from Train Datasets.


```{r}

missing_values <- eval_df %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Missing_Count")

print(missing_values)


```

### Removing Mostly NA and INDEX Column:

```{r}

eval_df <- eval_df[, !names(eval_df) %in% "INDEX"]
eval_df <- eval_df[, !names(eval_df) %in% "TEAM_BATTING_HBP"]
eval_df
```
### Mean/Median Imputation (For Numeric Data)

Instead of removing missing values, fill them with the mean or median:

```{r}

eval_df$TEAM_BATTING_SO[is.na(eval_df$TEAM_BATTING_SO)] <- mean(eval_df$TEAM_BATTING_SO, na.rm = TRUE)
eval_df$TEAM_BASERUN_SB[is.na(eval_df$TEAM_BASERUN_SB)] <- mean(eval_df$TEAM_BASERUN_SB, na.rm = TRUE)
eval_df$TEAM_BASERUN_CS[is.na(eval_df$TEAM_BASERUN_CS)] <- mean(eval_df$TEAM_BASERUN_CS, na.rm = TRUE)
eval_df$TEAM_FIELDING_DP[is.na(eval_df$TEAM_FIELDING_DP)] <- mean(eval_df$TEAM_FIELDING_DP, na.rm = TRUE)
eval_df$TEAM_PITCHING_SO[is.na(eval_df$TEAM_PITCHING_SO)] <- mean(eval_df$TEAM_PITCHING_SO, na.rm = TRUE)
```

#### 4. Verifying That Missing Values Are Fixed

```{r}

sum(is.na(eval_df))  # Total missing values
colSums(is.na(eval_df))  # Missing values per column

```

```{r}

skim(eval_df)
```

## Make Predictions Using the Improved_Model:

```{r}

# Generate predictions using the trained Improved Model
eval_df$PREDICTED_WINS <- predict(improved_model, newdata = eval_df)

# View a sample of predictions
head(eval_df$PREDICTED_WINS)

```
## Interpretation of Predicted TARGET_WINS Values

The predicted values represent the expected number of wins for teams based on their batting, pitching, and fielding statistics using the Improved Model.

For example:

A team with a predicted 69.15 wins is expected to win around 69 games in a full season.
A team with a predicted 82.52 wins is expected to perform better, likely finishing with around 82-83 wins.
The variation in predictions suggests that some teams are stronger than others based on key performance metrics (hits, home runs, walks, strikeouts, fielding errors, etc.).

### Evaluating the Accuracy of Predictions Against Actual TARGET_WINS

Now that we have predicted TARGET_WINS, we need to compare these predictions with the actual values in the TARGET_WINS column to assess model accuracy.

#### Step 1: Compute Error Metrics

```{r}
# Check structure of the dataset
str(eval_df)

# Check if ACTUAL_WINS and PREDICTED_WINS are numeric
is.numeric(eval_df$ACTUAL_WINS)
is.numeric(eval_df$PREDICTED_WINS)

```
#### Step 2: Add an INDEX Column to Each Dataset

```{r}
# Add a row index to both datasets
train_df$INDEX <- seq_len(nrow(train_df))  
eval_df$INDEX <- seq_len(nrow(eval_df))

```



#### Step 3: Align and Merge the Datasets

```{r}
# First, check if both datasets have a common identifier (like INDEX)
head(train_df$INDEX)
head(eval_df$INDEX)

# Merge train_df and eval_df using INDEX (if applicable)
merged_df <- merge(train_df[, c("INDEX", "TARGET_WINS")], 
                   eval_df[, c("INDEX", "PREDICTED_WINS")], 
                   by = "INDEX", all = FALSE)

# Now check if both columns have equal length
nrow(merged_df)  


```
Ensure data consistency before calculating accuracy metrics.
Convert to numeric to avoid errors in correlation or calculations.

#### Step 4: Compute Accuracy Metrics

```{r}
# Load necessary libraries
library(Metrics)  # For MAE and RMSE
library(ggplot2)  # For visualizations

# Compute MAE, RMSE, and R²
mae <- mae(merged_df$TARGET_WINS, merged_df$PREDICTED_WINS)  
rmse <- rmse(merged_df$TARGET_WINS, merged_df$PREDICTED_WINS)  
r_squared <- cor(merged_df$TARGET_WINS, merged_df$PREDICTED_WINS)^2  

# Print results
cat("Model Accuracy Metrics:\n")
cat("Mean Absolute Error (MAE):", round(mae, 2), "\n")
cat("Root Mean Squared Error (RMSE):", round(rmse, 2), "\n")
cat("R-Squared (R²):", round(r_squared, 4), "\n")

```
### Interpretation of Model Accuracy Metrics

The evaluation results indicate that our Improved Model is not performing well in predicting TARGET_WINS.

The Mean Absolute Error (MAE) of 13.87 suggests that, on average, predictions deviate by about 14 wins per team, which is quite high. 

Additionally, the Root Mean Squared Error (RMSE) of 17.38 implies that some predictions have even larger errors, highlighting potential inconsistencies or missing key predictors. 

Most concerning is the R-Squared (R²) value of 0.001, which means that the model explains almost none of the variation in team wins—essentially, the predictions are no better than random guesses. 

This suggests that the model may be overfitting the training data and failing to generalize, or that it lacks critical predictive features. 

To improve performance, we should consider adding more relevant predictors (such as ERA or OBP), removing unimportant or noisy variables, and potentially using alternative mo


### Applying Prediction to Our Evalution

With our models tested and evaluated, we can no apply our model to our final evaluation dataframe.

```{r final-predictions}

predictedWins = predict(stplg_model, eval_df)
eval_df["PREDICTED_WINS"] = predictedWins

```