---
title: "Assignment3_Data621"
author: "Mubashira Qari, Puja Roy"
date: "2025-03-09"
output: html_document
---

### Load Libraries

```{r, warning = FALSE, message = FALSE}
# Load required packages
library(htmltools)
library(caret)
library(pROC)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(corrplot)
library(skimr)
require(DataExplorer)
require(miscTools)
require(MASS)
require(performance)
require(lmtest)
require(mice)
require(glmnet)
require(Metrics) 
```

### Variables in the Dataset:

• zn: proportion of residential land zoned for large lots (over 25000 square feet) (predictor variable)  
• indus: proportion of non-retail business acres per suburb (predictor variable)  
• chas: a dummy var. for whether the suburb borders the Charles River (1) or not (0) (predictor variable)  
• nox: nitrogen oxides concentration (parts per 10 million) (predictor variable)  
• rm: average number of rooms per dwelling (predictor variable)  
• age: proportion of owner-occupied units built prior to 1940 (predictor variable)  
• dis: weighted mean of distances to five Boston employment centers (predictor variable)  
• rad: index of accessibility to radial highways (predictor variable)  
• tax: full-value property-tax rate per $10,000 (predictor variable)  
• ptratio: pupil-teacher ratio by town (predictor variable)  
• lstat: lower status of the population (percent) (predictor variable)  
• medv: median value of owner-occupied homes in $1000s (predictor variable)  
• target: whether the crime rate is above the median crime rate (1) or not (0) (response variable)  

###  load the dataset and understand its structure.

```{r}

crime_training_df <- read_csv("https://raw.githubusercontent.com/uzmabb182/Data_621/refs/heads/main/Assignment3/crime-training-data_modified.csv")

crime_evaluation_df <- read_csv("https://raw.githubusercontent.com/uzmabb182/Data_621/refs/heads/main/Assignment3/crime-evaluation-data_modified.csv")

head(crime_evaluation_df)



```


### Exploratory Data Analysis

```{r}
#dim(crime_training_df)
skim(crime_training_df)

### Puja Modified below:
### Load Data
crime_training_df <- read_csv("https://raw.githubusercontent.com/uzmabb182/Data_621/refs/heads/main/Assignment3/crime-training-data_modified.csv")

### Exploratory Data Analysis
# Summary Statistics
summary(crime_training_df)

# Correlation Heatmap
corr_matrix <- cor(crime_training_df %>% subset(select =-c(target)))
corrplot(corr_matrix, method = "color", tl.cex = 0.7)

# Boxplots for Key Variables
crime_training_df %>%
  gather(key = "Variable", value = "Value", -target) %>%
  ggplot(aes(x = as.factor(target), y = Value, fill = as.factor(target))) +
  geom_boxplot() +
  facet_wrap(~Variable, scales = "free") +
  theme_minimal()

### Data Preparation
# Log Transformations
crime_training_df <- crime_training_df %>%
  mutate(
    log_zn = log1p(zn),
    log_indus = log1p(indus),
    log_tax = log1p(tax),
    log_rad = log1p(rad),
    log_lstat = log1p(lstat),
    log_medv = log1p(medv)
  )

# Binning 'age' into Categories
crime_training_df <- crime_training_df %>%
  mutate(age_group = cut(age, breaks = c(0, 40, 70, 100), labels = c("Young", "Middle-aged", "Old")))

# Normalize 'dis'
crime_training_df <- crime_training_df %>%
  mutate(dis_scaled = scale(dis))

### Model Building
# Model 1: Baseline Logistic Regression
model1 <- glm(target ~ zn + indus + chas + nox + rm + age + dis + rad + tax + ptratio + lstat + medv, 
              data = crime_training_df, family = binomial)
summary(model1)

# Model 2: Stepwise Logistic Regression
model2 <- step(glm(target ~ ., data = crime_training_df, family = binomial), direction = "both")
summary(model2)

# Model 3: Logistic Regression with Transformed Variables
model3 <- glm(target ~ log_zn + log_indus + chas + nox + rm + age_group + dis_scaled + log_rad + log_tax + ptratio + log_lstat + log_medv, 
              data = crime_training_df, family = binomial)
summary(model3)

### Model Evaluation
# Compare AIC
AIC(model1, model2, model3)

# Confusion Matrix for Accuracy
pred1 <- ifelse(predict(model1, type = "response") > 0.5, 1, 0)
pred2 <- ifelse(predict(model2, type = "response") > 0.5, 1, 0)
pred3 <- ifelse(predict(model3, type = "response") > 0.5, 1, 0)

confusionMatrix(as.factor(pred1), as.factor(crime_training_df$target))
confusionMatrix(as.factor(pred2), as.factor(crime_training_df$target))
confusionMatrix(as.factor(pred3), as.factor(crime_training_df$target))

# ROC Curve Comparison
roc1 <- roc(crime_training_df$target, predict(model1, type = "response"))
roc2 <- roc(crime_training_df$target, predict(model2, type = "response"))
roc3 <- roc(crime_training_df$target, predict(model3, type = "response"))

plot(roc1, col = "red", main = "ROC Curves")
lines(roc2, col = "blue")
lines(roc3, col = "green")
legend("bottomright", legend = c("Model 1", "Model 2", "Model 3"), fill = c("red", "blue", "green"))

```



