---
title: "Data621_Final_Project"
author: "Mubashira Qari"
date: "2025-04-26"
output: html_document
---


### Load Libraries

```{r, warning = FALSE, message = FALSE}
# Load required packages
library(htmltools)
library(caret)
library(pROC)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(corrplot)
library(skimr)
require(DataExplorer)
require(miscTools)
require(MASS)
require(performance)
require(lmtest)
require(mice)
require(glmnet)
require(Metrics) 
library(patchwork)  # for combining ggplots
library(e1071)
library(car)
library(forcats)      # For better factor handling

```

###  load the dataset and understand its structure.

```{r, warning = FALSE, message = FALSE}
# echo=FALSE, include=FALSE

remote_work_df <- read_csv("https://raw.githubusercontent.com/uzmabb182/Data_621/refs/heads/main/Final_Project/Impact_of_Remote_Work_on_Mental_Health.csv")


head(remote_work_df)

```

### Checking & Cleaning Data

```{r}
glimpse(remote_work_df)
summary(remote_work_df)
skim(remote_work_df)
```

### Checking for missing values

```{r}

colSums(is.na(remote_work_df))

```
### Convert categorical variables to factors

```{r}

remote_work_df <- remote_work_df %>%
  mutate(across(where(is.character), as.factor))

```

### Checking Datatypes

```{r}
str(remote_work_df)
```
### Descriptive Statistics for Numeric Columns Summary

```{r}
remote_work_df %>%
  dplyr::select(where(is.numeric)) %>%
  summary()
```
### Descriptive Statistics for Categorical Columns Frequency

```{r}

remote_work_df %>%
  dplyr::select(where(is.factor)) %>%
  map(table)

```

### Distribution Plots for Age column

```{r}
ggplot(remote_work_df, aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(title = "Age Distribution", x = "Age", y = "Count")

```
### Distribution Plots for Work Location vs Stress Level

```{r}
ggplot(remote_work_df, aes(x = Work_Location, fill = Stress_Level)) +
  geom_bar(position = "dodge") +
  labs(title = "Stress Level by Work Location", x = "Work Location", y = "Count") +
  theme_minimal()

```
### Distribution Plots for Mental Health Condition by Gender

```{r}
ggplot(remote_work_df, aes(x = Gender, fill = Mental_Health_Condition)) +
  geom_bar(position = "fill") +
  labs(title = "Mental Health Condition Distribution by Gender", y = "Proportion") +
  theme_minimal() +
  coord_flip()

```
### Correlation (Numeric Features)

```{r}
# Correlation matrix
numeric_data <- remote_work_df %>%
  dplyr::select(where(is.numeric))

cor_matrix <- cor(numeric_data, use = "complete.obs")
round(cor_matrix, 2)

# Optional: Use corrplot for better visuals
library(corrplot)
corrplot(cor_matrix, method = "color", tl.cex = 0.8)

```
### Relationships Exploration for Productivity Change vs Mental Health

```{r}
ggplot(remote_work_df, aes(x = Mental_Health_Condition, fill = Productivity_Change)) +
  geom_bar(position = "dodge") +
  labs(title = "Productivity Change by Mental Health Condition") +
  theme_minimal()

```
###  Relationships Exploration for Sleep Quality by Region

```{r}
ggplot(remote_work_df, aes(x = Region, fill = Sleep_Quality)) +
  geom_bar(position = "dodge") +
  labs(title = "Sleep Quality Across Regions") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
### Automated Exploratory Report 

```{r}
# Generate an HTML report
create_report(remote_work_df, output_file = "eda_remote_work_report.html")

```
### Check for outliers or inconsistent entries


```{r}
# identify typos or categories like "Prefer not to say" that may need special handling
boxplot(remote_work_df$Hours_Worked_Per_Week, main = "Boxplot of Weekly Hours Worked")

# Check unique values in categorical columns
sapply(remote_work_df %>% dplyr::select(where(is.factor)), levels)

```
Possible Data Preparation Steps Outline (Post-Exploration)
1. Handle Missing Values
Check for NA values in both categorical and numerical columns.

Impute or drop:

For numerical features: use median or mean imputation.

For categorical features: use mode or a new category like "Unknown".

2. Encode Categorical Variables
Convert categorical features for ML models:

One-hot encoding: for unordered categories (e.g., Job_Role, Industry, Region).

Ordinal encoding: for ordered factors (e.g., Stress_Level, Satisfaction_with_Remote_Work).

3. Feature Engineering
Create new features if beneficial:

Binary flags (e.g., Has_Mental_Health_Issue = if Mental_Health_Condition ≠ "None")

Group rare categories (e.g., combine job roles or industries with low frequency)

Bucket Age or Years_of_Experience into groups if needed

4. Scale / Normalize Numeric Features
Normalize features like:

Hours_Worked_Per_Week

Social_Isolation_Rating

Work_Life_Balance_Rating

Use standardization (z-score) or min-max scaling depending on model choice

5. Balance the Dataset (if needed)
If target variable (e.g., Mental_Health_Condition or Productivity_Change) is imbalanced:

Consider SMOTE, undersampling, or class weighting for classification models

6. Remove Redundant Features
Drop non-informative or leakage-prone features:

Employee_ID (irrelevant for modeling)

Possibly Region or Work_Location if highly correlated with other variables

7. Outlier Detection
Use boxplots or z-scores to detect outliers in:

Hours_Worked_Per_Week, Social_Isolation_Rating, etc.

Decide whether to cap, transform, or remove

8. Ensure Consistent Factor Levels
Standardize labels for categorical values (e.g., "Prefer not to say" may need special handling)

Collapse categories for sparsity or interpretability

9. Create Target Variable (if modeling)
Define the dependent variable:

Classification: Mental_Health_Condition, Productivity_Change

Regression: could be derived scores (e.g., scale of mental health burden)

10. Train-Test Split (if modeling)
Split dataset (e.g., 70% training, 30% testing) using caret or rsample


