---
title: "Data621_Final_Project"
author: "Mubashira Qari"
date: "2025-04-26"
output:
  pdf_document: default
  html_document: default
---


### Load Libraries

```{r, warning = FALSE, message = FALSE}
# Load required packages
library(htmltools)
library(caret)
library(pROC)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(corrplot)
library(skimr)
require(DataExplorer)
require(miscTools)
require(MASS)
require(performance)
require(lmtest)
require(mice)
require(glmnet)
require(Metrics) 
library(patchwork)  # for combining ggplots
library(e1071)
library(car)
library(forcats)      # For better factor handling

```

###  load the dataset and understand its structure.

```{r, warning = FALSE, message = FALSE}
# echo=FALSE, include=FALSE

remote_work_df <- read_csv("https://raw.githubusercontent.com/uzmabb182/Data_621/refs/heads/main/Final_Project/Impact_of_Remote_Work_on_Mental_Health.csv")


head(remote_work_df)

```

### Checking & Cleaning Data

```{r}
glimpse(remote_work_df)
summary(remote_work_df)
skim(remote_work_df)
```

### Checking for missing values

```{r}

colSums(is.na(remote_work_df))

```
### Convert categorical variables to factors

```{r}

remote_work_df <- remote_work_df %>%
  mutate(across(where(is.character), as.factor))

```

### Checking Datatypes

```{r}
str(remote_work_df)
```
###  Dropping Employee_ID Column and Update df

```{r}
# Drop the Employee_ID column
remote_work_df <- remote_work_df %>% dplyr::select(-Employee_ID)

# Verify that the column is removed
glimpse(remote_work_df)


```

### Descriptive Statistics for Numeric Columns Summary

```{r}
remote_work_df %>%
  dplyr::select(where(is.numeric)) %>%
  summary()
```
### Descriptive Statistics for Categorical Columns Frequency

```{r}

remote_work_df %>%
  dplyr::select(where(is.factor)) %>%
  map(table)

```

### Distribution Plots for Age column

```{r}
ggplot(remote_work_df, aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(title = "Age Distribution", x = "Age", y = "Count")

```
### Distribution Plots for Work Location vs Stress Level

```{r}
ggplot(remote_work_df, aes(x = Work_Location, fill = Stress_Level)) +
  geom_bar(position = "dodge") +
  labs(title = "Stress Level by Work Location", x = "Work Location", y = "Count") +
  theme_minimal()

```
### Distribution Plots for Mental Health Condition by Gender

```{r}
ggplot(remote_work_df, aes(x = Gender, fill = Mental_Health_Condition)) +
  geom_bar(position = "fill") +
  labs(title = "Mental Health Condition Distribution by Gender", y = "Proportion") +
  theme_minimal() +
  coord_flip()

```
### Correlation (Numeric Features)

```{r}
# Correlation matrix
numeric_data <- remote_work_df %>%
  dplyr::select(where(is.numeric))

cor_matrix <- cor(numeric_data, use = "complete.obs")
round(cor_matrix, 2)

# Optional: Use corrplot for better visuals
library(corrplot)
corrplot(cor_matrix, method = "color", tl.cex = 0.8)

```
### Relationships Exploration for Productivity Change vs Mental Health

```{r}
ggplot(remote_work_df, aes(x = Mental_Health_Condition, fill = Productivity_Change)) +
  geom_bar(position = "dodge") +
  labs(title = "Productivity Change by Mental Health Condition") +
  theme_minimal()

```
###  Relationships Exploration for Sleep Quality by Region

```{r}
ggplot(remote_work_df, aes(x = Region, fill = Sleep_Quality)) +
  geom_bar(position = "dodge") +
  labs(title = "Sleep Quality Across Regions") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
### Automated Exploratory Report 

```{r}
# Generate an HTML report
create_report(remote_work_df, output_file = "eda_remote_work_report.html")

```
### Check for outliers or inconsistent entries


```{r}
# identify typos or categories like "Prefer not to say" that may need special handling
boxplot(remote_work_df$Hours_Worked_Per_Week, main = "Boxplot of Weekly Hours Worked")

# Check unique values in categorical columns
sapply(remote_work_df %>% dplyr::select(where(is.factor)), levels)

```
### Summary of Insights from Data Exploration

### Dataset Structure and Quality

The dataset includes a diverse set of variables like below: 

demographic (age, gender, region)
work-related (job role industry, work location) 
mental health (stress level mental health condition, access to resources)
and lifestyle factors (sleep quality, physical activity)

No missing values were reported — indicating a complete dataset suitable for direct analysis.

All categorical columns were successfully converted to factors, and numeric variables are in usable form.

Employee_ID was removed correctly, avoiding unnecessary noise.

### Interpretation of Charts:

### Age Distribution Chart

25–30, 40–50, and 50–55 age bins have the highest frequencies.

Each of these bins has 650–700 employees, indicating a strong middle-aged workforce presence.

The 18–22 group has the lowest count (under 150).

The 60+ group also has fewer participants (around 400), possibly due to retirements or reduced digital/remote work participation.

This is not a normal (bell-shaped) distribution. Instead, it’s closer to a uniform or flat distribution from ages 25 to 55.

The slight dip at the edges (youngest and oldest) is typical in workplace data where fewer very young or senior employees are present.

### Stress Level by Work Location Chart

Remote workers report the highest number of high stress levels, noticeably more than hybrid or onsite workers.

Onsite workers have the most balanced distribution, with slightly more reporting low stress than high.

Hybrid workers show a fairly even spread across all stress levels, suggesting a moderate stress profile.

### Mental Health Condition Distribution by Gender Chart

Anxiety is the most common condition across all genders, especially among males and females.

Non-binary and “Prefer not to say” groups have higher proportions of depression and burnout compared to the binary genders.

The "None" category (no mental health condition) is least prevalent in the non-binary and "Prefer not to say" groups.

Males have the highest proportion of "None" (no mental health condition).

### Correlation Matrix of Numeric Variables Chart

Strongest Positive Correlation:

Age and Years_of_Experience show a strong positive correlation (~0.9+), which is expected—older individuals typically have more work experience.

Moderate Positive Relationships:

Hours_Worked_Per_Week has a moderate positive correlation with Number_of_Virtual_Meetings, suggesting those who attend more meetings tend to work more hours.

Weak or No Correlations:

Most other relationships (e.g., Company_Support_for_Remote_Work, Work_Life_Balance_Rating, Social_Isolation_Rating) show low or negligible correlations, indicating they may vary independently.

No Evidence of Multicollinearity:

No pairs of variables (besides age/experience) show very high correlation (>0.85), so there's low risk of multicollinearity in predictive modeling.

### Productivity Change by Mental Health Condition Chart

Decrease in Productivity is most prominent among individuals with:

Depression (highest drop)

Burnout and Anxiety (also show high counts of decreased productivity)

Individuals with no mental health condition still report productivity decreases, but less frequently compared to those with depression.

Increase in productivity (green bars) is lowest among those with depression, suggesting a clear negative impact of depression on work output.

### Sleep Quality Across Regions Charts

Africa and Oceania have the highest counts of "Good" sleep quality, suggesting relatively better rest patterns among employees in those regions.

Asia and North America show a slightly higher proportion of "Poor" sleep quality, which may indicate higher stress, longer work hours, or less work-life balance.

Europe and South America present a balanced distribution, with no extreme dominance of any sleep quality level.

### Boxplot of Weekly Hours Worked Chart

Median (black line) is around 40 hours/week, which aligns with a standard full-time workload.

The interquartile range (IQR) spans approximately 30 to 50 hours, showing where most employees fall.

The minimum is around 20 hours, and the maximum is close to 60 hours.

There are no extreme outliers shown, but a few employees are working near the upper threshold, which may indicate potential overwork or burnout risk.


### Possible Data Preparation Steps Outline (Post-Exploration)

1. Handle Missing Values
Check for NA values in both categorical and numerical columns.

Impute or drop:

For numerical features: use median or mean imputation.

For categorical features: use mode or a new category like "Unknown".

2. Encode Categorical Variables
Convert categorical features for ML models:

One-hot encoding: for unordered categories (e.g., Job_Role, Industry, Region).

Ordinal encoding: for ordered factors (e.g., Stress_Level, Satisfaction_with_Remote_Work).

3. Feature Engineering
Create new features if beneficial:

Binary flags (e.g., Has_Mental_Health_Issue = if Mental_Health_Condition ≠ "None")

Group rare categories (e.g., combine job roles or industries with low frequency)

Bucket Age or Years_of_Experience into groups if needed

4. Scale / Normalize Numeric Features
Normalize features like:

Hours_Worked_Per_Week

Social_Isolation_Rating

Work_Life_Balance_Rating

Use standardization (z-score) or min-max scaling depending on model choice

5. Balance the Dataset (if needed)
If target variable (e.g., Mental_Health_Condition or Productivity_Change) is imbalanced:

Consider SMOTE, undersampling, or class weighting for classification models

6. Remove Redundant Features
Drop non-informative or leakage-prone features:

Employee_ID (irrelevant for modeling)

Possibly Region or Work_Location if highly correlated with other variables

7. Outlier Detection
Use boxplots or z-scores to detect outliers in:

Hours_Worked_Per_Week, Social_Isolation_Rating, etc.

Decide whether to cap, transform, or remove

8. Ensure Consistent Factor Levels
Standardize labels for categorical values (e.g., "Prefer not to say" may need special handling)

Collapse categories for sparsity or interpretability

9. Create Target Variable (if modeling)
Define the dependent variable:

Classification: Mental_Health_Condition, Productivity_Change

Regression: could be derived scores (e.g., scale of mental health burden)

10. Train-Test Split (if modeling)
Split dataset (e.g., 70% training, 30% testing) using caret or rsample


